---
title: "Week6hw"
output: html_document
author: "Your Name Here"
date: "`r Sys.Date()`"
---

Load Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(effectsize)
library(pwr)
library(emmeans)
```

## Problem 1: One-sample t-test

Dataset: mtcars
Question: Is the average miles per gallon (mpg) different from 20?

Response Variable: mpg (continuous)
Predictor: None (one group)
```{r}
#load data


# Visualize data & Check normality


# One-sample t-test


# Effect size


# Non-parametric alternative


# Plot Results




```

## Problem 2: Two-sample t-test

Dataset: iris
Question: Do setosa and versicolor differ in Sepal.Length?

Response Variable: Sepal.Length
Predictor: Species (categorical, 2 levels)

```{r}
#load data


#filter to include just two species 

# Plot a histogram & Check normality


# Two-sample t-test


# Effect size


# Non-parametric alternative


# Plot Results



```

## Problem 3: Chi-square test

Dataset: HairEyeColor
Question: Is hair color independent of eye color?

Response Variable: Eye color
Predictor Variable: Hair color

```{r}
#load the dataset 


#create a contingency table with counts for each hair/eye color combination
#you'll need to combine the counts across sex in order to just use hair and eye color (group_by and summarise)
#you might want to use pivot_wider to get the levels of eye color as rows and levels of hair color as columns (or vice versa) and column_to_rownames() to make it a proper contingency table



# Chi-square test


# Observed and expected counts


# Plot



```

## Problem 4: One-way ANOVA & correlation

Dataset: mtcars
Question 4a (ANOVA): Does the number of cylinders (cyl) predict horsepower (hp)?
Question 4b (Correlation): Is wt correlated with mpg?
```{r}
#mtcars is already loaded 

#Look at the data and check assumptions 


# ANOVA


# Post-hoc (where are the differences?)


# Plot



# Check correlation assumptions 


# Correlation


# Spearman if assumptions violated

# Plot


```

