---
title: "Week3inClass"
author: "Lily Johnson-Ulrich"
date: "2025-09-05"
output: html_document
---

## Step 1. Load your libraries and the the dataframe 'iris' as a tibble (with a new or the same name)
```{r}

```

## Step 2: Data Types

1. Look at the dataset with ?iris and and by typing your table name. 
2. Identify which variables are qualitative nominal, ordinal, discrete quantitative, and continuous quantitative. Use a function to examine each variable and note in the code chunk using a # which variables are which. Convert any to the correct data type if they are labeled incorrectly. 

```{r}
#1. 


#2. 



```


## Step 3: Visualizing Data

### A. Continuous Variables
Make a plot that shows the frequency of various scores of *one* of your continuous variables. 

```{r}
#Create your frequency plots here



```

Create a histogram and experiment with binwidth (using a different continuous variable than above).
What happens if you make binwidth larger? Smaller? Show your code to demonstrate what happens! 
```{r}
#Plot your histograms here. 


```

### B. Relationship Between Two Continuous Variables

Make a scatterplot with two continuous variables. 
```{r}



```

### C. Categorical Variable

Make a bar plot showing the number of flowers in each Species.
Are the sample sizes balanced across species?
```{r}



```


## Step 4: Summary Statistics: Describing your Data

Compute summary statistics of Petal.Length for each species.
```{r}
#1. Create a density plot for Petal Length 


#2. Calculate summary stats for Petal Length and use geom_vline() to add verticle lines for the mean, median, and mode 


# How would you describe shape of this distribution in terms of a) symmetry, b) uniformity, and c) variation? 



# What might be affecting b in particular? 
# Hint: Create a grouped density plot by species 



#which species has the longest petal.length on average? 

```

## Step 5: Plotting Summary Stats
Make a bar chart of mean Sepal.Length by species with ±SE error bars. Hint: save a table with summary stats. 
```{r}




#why might we choose SE here instead of SD? 

```