---
title: "Dog Bite Data: GLMM & Zero-Inflation Homework"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(glmmTMB)
library(DHARMa)
```

## Introduction

In this homework, we explore dog bite counts in a dataset with multiple breeds. Dogs are observed over multiple months, and the number of bites is recorded. Some dogs never bite, making the data zero-inflated. We’ll practice GLMMs, zero-inflation, and random effects.

Simulate dataset:
```{r}
set.seed(42)

n_dogs <- 150
n_breeds <- 15
n_months <- 6

# Dog-level info

dog_df <- data.frame(
dogID = paste0("Dog", 1:n_dogs),
breed = sample(paste0("Breed", 1:n_breeds), n_dogs, replace = TRUE),
age_cat = sample(c("puppy","adult","senior"), n_dogs, replace = TRUE),
sex = sample(c("M","F"), n_dogs, replace = TRUE),
trained_commands = sample(5:25, n_dogs, replace = TRUE)
)

# Expand to repeated measures

dog_obs <- dog_df %>%
slice(rep(1:n(), each = n_months)) %>%
mutate(month = rep(1:n_months, times = n_dogs))

# Random effects

breed_re <- rnorm(n_breeds, 0, 0.7)
names(breed_re) <- paste0("Breed", 1:n_breeds)
dog_re <- rnorm(n_dogs, 0, 0.5)
names(dog_re) <- dog_df$dogID

# Fixed effects

age_eff <- c("puppy"=0.5, "adult"=0, "senior"=-0.3)
train_eff <- -0.05  # more trained commands, fewer bites

# Linear predictor

dog_obs <- dog_obs %>%
rowwise() %>%
mutate(
log_lambda = -1 +
age_eff[age_cat] +
train_eff * trained_commands +
breed_re[breed] +
dog_re[dogID],
lambda = exp(log_lambda),
# zero-inflated Poisson
bite_count = ifelse(runif(1) < 0.4, 0, rpois(1, lambda))
) %>%
ungroup()

head(dog_obs)

```

## Problem 1: Data Exploration

1. Make a histogram of bite_count. What do you notice about zeros? 
[Type answer here.]

2. Make a boxplot of bite_count by age category. Any trends?
[Type answer here.]


```{r}
# Histogram

ggplot(dog_obs, aes(x = bite_count)) +
geom_histogram(binwidth = 1, fill="lightblue", color="black") +
labs(x="Bite count", y="Number of observations")

# Boxplot

ggplot(dog_obs, aes(x = age_cat, y = bite_count)) +
geom_boxplot(fill="lightgreen") +
labs(x="Age category", y="Bite count")

```

**Question:** Based on the histogram, do you think a Poisson model is sufficient? Why or why not?
[Insert Answer Here]

## Problem 2: Fit a GLMM

Fit a Poisson GLMM predicting bite_count from trained_commands and age_cat, with breed as a random effect.
```{r}
# Fill in the code

# Hint: family = poisson, random effect = (1|breed)

m1 <- glmmTMB( # your formula here
  family = poisson,
  data = dog_obs)

summary(m1)
icc(m1)

emmeans(m1, ~age_cat, type="response")
estimate_slopes(m1, trend="trained_commands")
```

**Questions:**

Interpret the fixed effect of trained_commands. 
[Type answer here.]

How much variation is there between breeds (look at random effect variance)?
[Type answer here.]

What might be a problem with this model given the histogram of zeros?
[Type answer here.]

## Problem 3: Residual diagnostics
```{r}
#Simulate residuals using DHARMa:
res <- simulateResiduals(m1)
plot(res)

#Check overdispersion
testDispersion(res)
```

**Questions:**

Does the QQ plot suggest any problems?
[Type answer here.]

Is there evidence of overdispersion?
[Type answer here.]

Based on these checks, what model adjustment might you consider? (Hint: zero-inflated Poisson, additional random effects, or both)
[Type answer here.]

## Problem 4: Fit a zero-inflated poisson.
```{r}
m2 <- glmmTMB(
  bite_count ~ trained_commands + age_cat + (1|breed),
  ziformula = ~1,
  family = poisson,
  data = dog_obs
)
summary(m2)
res2 <- simulateResiduals(m2)
plot(res2)
```

**Question**: Is the zero-inflated model a better fit? 
[Type answer here.]

# Wrap-up

1. Knit your file to .html
2. Put both files in your git repo folder 
3. Use 'git pull' to download any changes from github.com
4. Use 'git add filename' to add them to git tracking
5. Use 'git commit -m "comment"' to commit them for upload
6. Use 'git push' to upload your changes from github.com 
